%
% Sjabloon voor de Master of Science in de ingenieurswetenschappen: computerwetenschappen
%
%%%%%%%%%%% Wijzig niets boven deze regel %%%%%%%%%%
% Wijzig hieronder eventueel de optie
% ai -> hoofdoptie Artificiële intelligentie
% ci -> hoofdoptie Computationele informatica
% gs -> hoofdoptie Gedistribueerde systemen
% mmc -> hoofdoptie Mens-machine communicatie
% se -> hoofdoptie Software engineering
% vs -> hoofdoptie Veilige software
\documentclass[master=cws,masteroption=ai]{kulemt}
%
% Vul de titel van jouw masterproef hieronder in tussen { en }.
\setup{title={Algebraic Subtyping for Algebraic Effects and Handlers},
%
% Vul hieronder namen in, steeds Voornaam Naam.
% Indien meerdere auteurs, assessoren, assistenten, scheidt hun namen
% met "\and".
author={Axel Faes},
promotor={Prof.\,dr.\,ir.\ Tom Schrijvers},
assessor={Amr Hany Shehata Saleh\\Prof.\,dr. Bart Jacobs},
assistant={Amr Hany Shehata Saleh}}
%
\setup{filingcard, % Deze regel niet wijzigen
%
% Vul de vertaalde titel van jouw masterproef hieronder in tussen { en }.
translatedtitle={Algebraic Subtyping for Algebraic Effects and Handlers},
%
% UDC nummer is richtingafhankelijk. zie http://www.udcsummary.info/php/index.php voor het UDC nummer.
udc=621.3,
%
% Hieronder, tussen { en } een korte samenvatting toevoegen.
% Lege regels en het commando \par zijn niet toegelaten.
% Wees voorzichtig met speciale TeX-tekens #$%&^_~{}\ !!
shortabstract={Algebraic effects and handlers benefit from a custom type-\&-effect system, a type system that also tracks which effects can happen in a program. Several such type-\&-effect systems have been proposed in the literature, but all are unsatisfactory. Recently, Stephen Dolan (University of Cambridge, UK) presented a novel type system that combines subtyping and parametric polymorphism in a particulary attractive and elegant fashion. A cornerstone of his design are the algebraic properties that the subtyping relation should respect. In this work, a type-\&-effect system is derived that extends Dolan's elegant type system with effect information. This type-\&-effect system inherits Dolan's harmonious combination of subtyping (in our case induced by a lattice structure on the effect information) with parametric polymorphism and preserves all of its desirable properties (both low-level algebraic properties and high-level meta-theoretical properties like type soundness and the existence of principal types). This type-\&-effect system has been implemented in the \eff programming language in order to provide a proof-of-concept.}}
%%%%%%%%% Wijzig niets onder deze regel %%%%%%%%
\setup{coverpageonly}
\begin{document}
\end{document}
