\documentclass[master=ecws, masteroption=ai]{kulemt} %master=cws
\setup{title={Algebraic Subtyping for Algebraic Types and Effects},
  author={Axel Faes},
  promotor={Prof.\,dr.\,ir.\ Tom Schrijvers},
  assessor={Amr Hany Shehata Saleh, Prof.\,dr.\,ir.\ Bart Jacobs},
  assistant={Amr Hany Shehata Saleh}}
% The following \setup may be removed entirely if no filing card is wanted
\setup{filingcard,
  translatedtitle=,
  udc=,
  shortabstract={}}

% Choose the main text font (e.g., Latin Modern)
\setup{font=lm}

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\usepackage{array}
\usepackage{mathpartir}
\usepackage{xspace}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{newtxmath}

%% Tikz Needed packages
\usepackage{pgfplots}
\pgfplotsset{width=7cm,compat=1.8}
\usepackage{pgfplotstable}
\renewcommand*{\familydefault}{\sfdefault}

% Finally the hyperref package is used for pdf files.
% This can be commented out for printed versions.
\usepackage[pdfusetitle,colorlinks,plainpages=false]{hyperref}

\newcommand{\C}{\underline{C}}
\input{../spec/tex/00-macros}
\newcommand{\todo}[1]{\textcolor{red}{\textsc{Todo:} #1}}

% Needed to use the files from spec
\let\subparagraph\paragraph
\let\paragraph\subsubsection
\let\subsubsection\subsection
\let\subsection\section
\let\section\chapter

\begin{document}

\begin{preface}
  I would like to thank everybody who kept me busy and supported me the last year,
  especially my promoter and my assistants. I would also like to thank the
  jury for reading the text.
\end{preface}

\tableofcontents*

\begin{abstract}
  Algebraic effects and handlers benefit from a custom type-\&-effect system, a type system that also tracks which effects can happen in a program. Several such type-\&-effect systems have been proposed in the literature, but all are unsatisfactory. Recently, Stephen Dolan (University of Cambridge, UK) presented a novel type system that combines subtyping and parametric polymorphism in a particulary attractive and elegant fashion. A cornerstone of his design are the algebraic properties that the subtyping relation should respect. In this work, a type-\&-effect system is derived that extends Dolan's elegant type system with effect information. This type-\&-effect system inherits Dolan's harmonious combination of subtyping (in our case induced by a lattice structure on the effect information) with parametric polymorphism and preserves all of its desirable properties (both low-level algebraic properties and high-level meta-theoretical properties like type soundness and the existence of principal types).
\end{abstract}

\listoffigures
\listoftables

% Now comes the main text
\mainmatter

\section{Introduction}
\input{../spec/tex/01-introduction}

\section{Simply Typed Lambda Calculus}
\input{../spec/tex/02-background}

\section{Algebraic effects and handlers (\eff)}
\input{../spec/tex/04-eff-intro}
\input{../spec/tex/05-eff-types}
\input{../spec/tex/06-eff-type-system}

\section{Core Language (\core)}
\input{../spec/tex/07-core-intro}
\input{../spec/tex/08-core-types}
\input{../spec/tex/09-core-type-system}
% \input{../spec/tex/10-core-subtyping}
\input{../spec/tex/11-core-typing-rules}
\input{../spec/tex/12-core-reformulated-typing}

\section{Proofs}
\input{../spec/tex/13-proofs}

\section{Type Inference}
% \input{../spec/tex/14-elaboration}
% \input{../spec/tex/15-constraints}
\input{../spec/tex/16-polarity}
\input{../spec/tex/17-unification}
\input{../spec/tex/18-type-inference}

\section{Implementation}
\input{../spec/tex/19-implementation}

\section{Evaluation}
\input{../spec/tex/20-evaluation}

\section{Conclusion}
\input{../spec/tex/21-conclusion}

\appendixpage*          % if wanted
\appendix
% \chapter{Appendix A}
% \input{../spec/tex/appendix}

\backmatter

% The bibliography comes after the appendices.
% You can replace the standard "abbrv" bibliography style by another one.
\bibliographystyle{abbrv}
\bibliography{../spec/bib/main}
\nocite{*}

\end{document}
